A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-05 paper.tex
Further notes on uncertainty

The necessity of complete precision about all recombination
events under the eARG encoding has been
conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
For example, \citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events''. While there are certainly cases
when inferring such fully precise ARGs is useful, for example when
studying the details of a small number of recombination
events~\citep{rasmussen2022espalier}, it is worth considering
whether sufficient information exists to support such inferences
and indeed whether this precision is required, or useful,
for downstream applications.

The amount of information about recombination that is inferrable
is the subject of much research
has a long history [summarise this in a couple of sentences?]
[Cite: \citep{myers2002detection,hayman2023recoverability}
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely},
% TODO look this up. I expect it's basically the same??
and the block-based ARG data structure of
\citep{palamara2016argon}),
here we consider the unique sets of
samples that coalesce at a node in the ARG over a particular
genome interval, and the fundamental limitations they place
on ARG inference.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully precise ARGs
from data,
we should also consider the value that such precise estimates provide
to downstream applications.
To date, few downstream applications using estimated ARGs
make use of such complete recombination information.
The \relate\ selection statistic~\citep{speidel2019method}
computes a $p$-value for selection [FIXME] for each
local tree independently by comparing the rates of coalescence
at various time-points.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in a ``full'' ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
[MORE applications]
The main application for fully precise ARGs thus far has been
to be to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

>>>>

Uncertainty is a central element of inference about recombination,
and, even with small samples, we can only expect to uniquely identify
lineages and breakpoints in special cases.
[summarise info inferable about recomb in few sentenances here?
Cite: \citep{myers2002detection,hayman2023recoverability}]
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
The fact that the classical ``event'' ARG (eARG) encoding
\emph{requires} completely precise estimates about recombination
is therefore a fundamental limitation, and a central message
of this paper.

% The necessity of complete precision about all recombination
% events under the eARG encoding has been
% conflated with the idea of a ``true'' or ``full'' ARG. [Cites]
% For example, \citet{brandt2021evaluation} define a ``full'' ARG as
% ``a structure that encodes all
% coalescence and recombination events''. While there are certainly cases
% when inferring such fully precise ARGs is useful, for example when
% studying the details of a small number of recombination
% events~\citep{rasmussen2022espalier}, it is worth considering
% whether sufficient information exists to support such inferences
% and indeed whether this precision is required, or useful,
% for downstream applications.

% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
% (equivalent to the ``bricked tree sequence'' of
% \cite{nowbandegani2023extremely},
% % TODO look this up. I expect it's basically the same??
% and the block-based ARG data structure of
% \citep{palamara2016argon}),
% here we consider the unique sets of
% samples that coalesce at a node in the ARG over a particular
% genome interval, and the fundamental limitations they place
% on ARG inference.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
to downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection statistic~\citep{speidel2019method}
computes selection $p$-values by comparing the rates of coalescence
at various time-points.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to be to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.
